---
title: "Script 01 - Drivers of FKBP5 DNA methylation: sample demographics"
author: "Vera N. Karlbauer"
contact: vera_karlbauer@psych.mpg.de
date: "2024-07-25"
output: 
  html_document:
  toc: true
  code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 1 for manuscript on Drivers of FKBP5 DNA methylation.
- Create table of sample demographics and distribution of key variables (table 1).
- T-test group differences between maltreated and non-maltreated children on key variables.

# Setup

### Step 0: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(haven)
library(formattable)
library(tidyr)

```

### Step 1: Import data

```{r}
# all processed data over time 
load("../../01_data/02_proc/all_time.rda")

# additional phenotypes
psych_pheno <- read_sav("../../01_data/01_raw/LCS_T0T4_TP2_Feb2019_fullversion_long-2.sav")

```

### Step 2: Process data

```{r}
## phenotypes
# select relevant variables
psych_pheno <- select(psych_pheno,
                      SUBJECT,
                      VISIT,
                      SES
                      )
# delete labels
psych_pheno <- zap_labels(psych_pheno)

# subject as character
psych_pheno$SUBJECT <- as.character(psych_pheno$SUBJECT)

## covariates
# select relevant variables  
covariates <- select(all_time,
                     Subject,
                     Case.Control,
                     age,
                     Sex,
                     Time,
                     PAPA_any,
                     genotype,
                     Cort_AUCg,
                     CRP_Ln,
                     CD34,
                     CD14,
                     Buccal)

```


### Step 3: Create joint dataframe

```{r}
# join dataframes
covariates <- left_join(covariates, psych_pheno, by = c("Subject" = "SUBJECT", "Time" = "VISIT"))
covariates_t0 <- filter(covariates, Time == 0)
covariates_t2 <- filter(covariates, Time == 2)

```

# Create demographics table

```{r}
# get total N per time point
table(covariates$Time)

# maltreatment frequency and % per time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(cnt = n()) %>%
  mutate(freq = formattable::percent(cnt / sum(cnt), digits = 1))

# sex per maltreatment and time point
covariates %>%
  group_by(Time, Case.Control, Sex) %>%
  summarise(cnt = n())

# age and sd per maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_age = round(mean(age), digits = 2), sd_age = round(sd(age), digits = 2))

# genotype frequency and % per time point
covariates %>%
  group_by(Time, genotype) %>%
  summarise(cnt = n()) %>%
  mutate(freq = formattable::percent(cnt / sum(cnt), digits = 1))

# SES and sd per maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_ses = round(mean(SES), digits = 2), sd_ses = round(sd(SES), digits = 2))

# psychiatric diagnosis and % by maltreatment and time point
covariates %>%
  drop_na(PAPA_any) %>%
  group_by(Time, Case.Control, PAPA_any) %>%
  summarise(cnt = n()) %>%
  mutate(freq = formattable::percent(cnt / sum(cnt), digits = 1))

# cort levels by maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_cort = round(mean(Cort_AUCg, na.rm = TRUE), digits = 2), sd_cort = round(sd(age), digits = 2))

# CRP levels by maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_crp = round(mean(CRP_Ln, na.rm = TRUE), digits = 2), sd_crp = round(sd(CRP_Ln, na.rm = TRUE), digits = 2))

# CD34 proportion (%) by maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_cd34 = round(mean(CD34, na.rm = TRUE), digits = 2), sd_cd34 = round(sd(CD34, na.rm = TRUE), digits = 2))

# CD14 proportion (%) by maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_cd14 = round(mean(CD14, na.rm = TRUE), digits = 2), sd_cd14 = round(sd(CD14, na.rm = TRUE), digits = 2))

# Buccal proportion (%) by maltreatment and time point
covariates %>%
  group_by(Time, Case.Control) %>%
  summarise(mean_buccal = round(mean(Buccal, na.rm = TRUE), digits = 2), sd_buccal = round(sd(Buccal, na.rm = TRUE), digits = 2))

```

# Test group differences maltreated / non-maltreated

## Sex

```{r}
# Fisher's exact test T0
stats::chisq.test(covariates_t0$Sex, covariates_t0$Case.Control)

# Fisher's exact test T2
stats::chisq.test(covariates_t2$Sex, covariates_t2$Case.Control)

```

## Age

```{r}
# t-test T0
t.test(age ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(age ~ Case.Control, data = covariates_t2)

```

## SES

```{r}
# T-test
# t-test T0
t.test(SES ~ Case.Control, data = covariates_t0)
format(2.2e-16, scientific = FALSE)

# t-test T2
t.test(SES ~ Case.Control, data = covariates_t2)
format(1.828e-14, scientific = FALSE)

```

## Psychiatric diagnosis

```{r}
# Fisher's exact test T0
stats::chisq.test(covariates_t0$PAPA_any, covariates_t0$Case.Control)
format(9.198e-07, scientific = FALSE)

# Fisher's exact test T2
stats::chisq.test(covariates_t2$PAPA_any, covariates_t2$Case.Control)
format(1.771e-06, scientific = FALSE)

```

## Cortisol

```{r}
# t-test T0
t.test(Cort_AUCg ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(Cort_AUCg ~ Case.Control, data = covariates_t2)

```

## CRP

```{r}
# t-test T0
t.test(CRP_Ln ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(CRP_Ln ~ Case.Control, data = covariates_t2)

```

## CD34

```{r}
# t-test T0
t.test(CD34 ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(CD34 ~ Case.Control, data = covariates_t2)

```

## CD14

```{r}
# t-test T0
t.test(CD14 ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(CD14 ~ Case.Control, data = covariates_t2)

```


## Buccal

```{r}
# t-test T0
t.test(Buccal ~ Case.Control, data = covariates_t0)

# t-test T2
t.test(Buccal ~ Case.Control, data = covariates_t2)

```

## Genotype

```{r}
# chisq T0
chisq.test(covariates_t0$genotype, covariates_t0$Case.Control)

# chisq test T2
chisq.test(covariates_t2$genotype, covariates_t2$Case.Control)

```
