---
title: "Variance Partitioning"
author: "Vera Natalia Karlbauer"
contact: vera_karlbauer@psych.mpg.de
date: "2024-08-02"
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 4 for manuscript on Drivers of FKBP5 DNA methylation.
- Prepares data on FKBP5 methylation & covariates for variance partitioning.
- Runs variance partitioning analysis to determine drivers of FKBP5 methylation & exports results.
- Plots results as horizontal barplot (figure 2).

# Setup

## Step 1: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(ggplot2)
library(variancePartition)
library(tidyr)
library(stringr)
library(colorBlindness)
library(haven)
library(janitor)
library(tibble)
library(writexl)
library(scales)

```

## Step 2: Load data

```{r}
# set temporary wd for chunk to folder with preprocessed data
setwd("../../01_data/02_proc")

# all_time = methylation & covariates for T0 & T2
load("all_time.rda")

# additional phenotypes
psych_pheno <- read_sav("../01_raw/LCS_T0T4_TP2_Feb2019_fullversion_long-2.sav")
# select relevant variables
psych_pheno <- select(psych_pheno,
                      SUBJECT,
                      VISIT,
                      SES
                      )
# delete labels
psych_pheno <- zap_labels(psych_pheno)
# subject as character
psych_pheno$SUBJECT <- as.character(psych_pheno$SUBJECT)

```

## Step 3: Remove missing values & prepare data

```{r}
### methylation over time
# join additional phenotypes
all_time <- left_join(all_time, psych_pheno, by = c("Subject" = "SUBJECT", "Time" = "VISIT"))
# remove CPGs with large amounts of missings
all_time_nomiss <- dplyr::select(all_time, - c(CTCF_PCR_5.1_35704149, CTCF_PCR_5.1_35704224, FKBP5_PCR_12_35683488, FKBP5_PCR_18_35697759, FKBP5_PCR_18_35697810, FKBP5_int5_PCR_3_35578830, FKBP5_int5_PCR_3_35578739, FKBP5_int5_PCR_3_35578891, FKBP5_int5_PCR_2_35569751))
# remove unnecessary variables
all_time_nomiss <- dplyr::select(all_time_nomiss, 
                                 -c(mergeID, MethID, cell_proportions, PAPA_sum, AA_AUCg_FR_or, CRP_Ln_or, 
                                    Cort_AUCg_or, Cort_inter_AUCg, Cort_baseline, AA_AUCg_FR, life_events_A_total, 
                                    life_events_A_cat, life_events_B_total, life_events_B_cat, adversity_score,
                                    maltreatment_intensity, Nr_external, Nr_internal, preg_precl, preg_csection,
                                    preg_preterm))
# reduce to complete cases & set sample ID as rownames
all_time_nomiss <- all_time_nomiss %>%
  drop_na %>%
  column_to_rownames(var = "Sample_ID")
# count nr of distinct subjects 
all_time_nomiss %>%
  distinct(Subject) %>%
  nrow()

# create methylation matrix
methylation_time_nomiss <- select(all_time_nomiss, contains("PCR"))
methylation_time_nomiss <- t(methylation_time_nomiss)

```

# Analysis

## Step 1: Variance partitioning over time

```{r}
# define formula (full model with all factors)
form <- ~ Time + age + Sex + Case.Control + PAPA_any + genotype + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score + gc_exposure_score + Cort_AUCg + CRP_Ln + biomom_psych_diag +
  life_events_total + preg_complications + SES + (1 | Subject)

# variance partition
varPart_time <- fitExtractVarPartModel(exprObj = methylation_time_nomiss, formula = form, data = all_time_nomiss, useWeights = FALSE, showWarnings = TRUE, quiet = FALSE)
# create sorted results object
varPart_time_sorted <- sortCols(varPart_time)
# custom order of CpGs according to position
cpg_order <- as.data.frame(matrix(NA, ncol = 0, nrow = 39))
cpg_order$cpg <- rownames(varPart_time_sorted)
cpg_order$position <- as.numeric(str_extract(cpg_order$cpg, "([3]{1}[0-9]{7})"))
cpg_order <- cpg_order[order(cpg_order$position),]
varPart_time_sorted <- varPart_time_sorted[order(factor(rownames(varPart_time_sorted), 
                                                        levels = cpg_order$cpg)),]
print(varPart_time_sorted)

# get range of total variance explained
round(range(rowSums(select(varPart_time_sorted, -Residuals))) * 100, digits = 2)
rowSums(select(varPart_time_sorted, -Residuals))

# get % of variance explained by cell counts
var_cellcounts <- varPart_time_sorted$CD34 + varPart_time_sorted$CD14 + varPart_time_sorted$Buccal
round(range(var_cellcounts) * 100, digits = 2)

# get % of variance explained by CD34 cells
round(range(varPart_time_sorted$CD34) * 100, digits = 3)

# get % of variance explained by buccal cells
round(range(varPart_time_sorted$Buccal) * 100, digits = 3)

# get % of variance explained by CD14 cells
round(range(varPart_time_sorted$CD14) * 100, digits = 3)

# get % of variance explained by prenatal exposures
var_exposures <- varPart_time_sorted$gc_exposure_score + varPart_time_sorted$alcohol_exposure_score + varPart_time_sorted$smoke_exposure_score
round(range(var_exposures) * 100, digits = 2)

# get % of variance explained by GC exposure
round(range(varPart_time_sorted$gc_exposure_score) * 100, digits = 3)

# get % of variance explained by alc exposure
round(range(varPart_time_sorted$alcohol_exposure_score) * 100, digits = 3)

# get % of variance explained by smoke exposure
round(range(varPart_time_sorted$smoke_exposure_score) * 100, digits = 3)

# get % of variance explained by maltreatment
round(range(varPart_time_sorted$Case.Control) * 100, digits = 2)

# get top 5 CpGs for variance explained by maltreatment
varPart_time_sorted %>%
  slice_max(Case.Control, n = 5) %>%
  select(Case.Control)

# export results
# export as .txt
write.table(varPart_time_sorted, file = "../03_results/variance_partitioning/results_varPart_time.txt", sep = "\t")
# convert to %
varPart_time_sorted_perc <- varPart_time_sorted
varPart_time_sorted_perc[1:23] <- sapply(varPart_time_sorted_perc[1:23], function(x) percent(x, accuracy=0.001))
varPart_time_sorted_perc <- varPart_time_sorted_perc %>%
  rownames_to_column(var = "CpG") %>%
  relocate(CpG, .before = CD34)
# export as xlsx for manuscript
write_xlsx(varPart_time_sorted_perc, path = "../05_tables/variance_partitioning/results_varPart_time.xlsx", col_names = TRUE)
```


## Step 2: Plot results

```{r}
# generate colors
SteppedSequential5Steps
Green2Magenta16Steps

color_exposures <- c("#990F0F",  "#CC5151", "#E57E7E")
# gc score, alcohol score, smoking score
  
color_celltypes <- c("#0F6B99","#51A3CC", "#7EC3E5")
# CD34, buccal, CD14

color_geno <- c("#6B990F","#85B22C", "#A3CC51", "#C3E57E")
# C1, C2, genotype, C3

color_pheno <- c("#99540F", "#B26F2C", "#CC8E51", "#E5B17E", "#FFD8B2")
# biomom psych diag, maltreatment, preg complications, psych diagnosis, life events

color_biodata <- c("#860086", "#F100F1")
# cort, CRP
  
color_demo <- c("#260F99", "#422CB2", "#6551CC")
# SES, sex, age

color_time <- "#8F7EE5"

color_subject <- "#BFB2FF"

color_res <- "#D3D3D3"

# set vector with distinguishable color values
color_values <- c(color_celltypes[1], color_exposures[1], color_celltypes[2], color_exposures[2], color_demo[1],
                  color_exposures[3], color_pheno[1], color_pheno[2], color_demo[2], color_demo[3],
                  color_geno[1], color_pheno[3], color_celltypes[3], color_geno[2], color_geno[3], color_time,
                  color_biodata[1], color_pheno[4], color_biodata[2], color_geno[4], color_pheno[5],
                  color_subject, color_res)

# check colorblind friendliness
displayAllColors(color_values, color="white")

plotPercentBars(varPart_time_sorted) +
  theme(plot.margin = margin(1,1,1,1, "cm")) + 
  theme(axis.text.y = element_text(size = 7)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(legend.key.size = unit(0.7, 'cm')) +
  scale_fill_manual(values = color_values, labels = c("CD34 cells", "Glucocorticoid exposure score", "Buccal cells", "Prenatal alcohol exposure score", "SES", "Prenatal smoke exposure score", "Maternal psychopathology", "Maltreatment", "Sex", "Age", "C1 genetic component", "Pregnancy complications", "CD14 cells", "C2 genetic component", "rs1360780 genotype", "Time", "Cortisol", "Psychiatric diagnosis", "CRP", "C3 genetic component", "Life events", "Subject", "Residuals"))
                                                      
ggsave(filename = "figure_2_varpart_time_barplot.png", path = "../04_figures", device = 'png', width = 10, height = 5, dpi=700)

```

