---
title: 'Prenatal GC exposure: Correlations'
author: "Vera N. Karlbauer"
date: "`r Sys.Date()`"
contact: vera_karlbauer@psych.mpg.de
date: 02-08-2024
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 9 for manuscript on Drivers of FKBP5 DNA methylation.
- Analyzes associations of prenatal glucocorticoid exposure with different maternal and childhood phenotypes. 
- Runs linear regression models for T0 and T2. 
- Exports results and publication tables (table 2).  
- Phenotypes of interest: 
  - maternal mental disorder
  - self-reported pregnancy complications 
  - pre-term birth 
  - maltreatment (categorical)
  - maltreatment severity (continuous)
  - stressful life events & contextual stressors experienced by children.
- Prenatal glucocortocid exposure was calculated according to Provencal et al. (2019): https://doi.org/10.1073/pnas.1820842116, supplement S8. 


# Setup

## Step 0: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(lme4)
library(stringr)
library(tibble)
library(rsq)

```


## Step 1: Load data

```{r, warning = FALSE}
# set temporary wd for chunk to folder with preprocessed data
setwd("../../01_data/02_proc")

# all_t0 = methylation & covariates for T0
load("all_t0.rda")

# all_t2 = methylation & covariates for T2
load("all_t2.rda")

# create empty summary table
sumtable_gc <- data.frame(matrix(nrow = 0, ncol = 5))
colnames(sumtable_gc) <- c("phenotype", "time", "coef", "p", "adj_partial_Rsquared")

```

# Analysis
## GC exposure & maternal mental disorder

### T0

```{r}
# get distributions
table(all_t0$biomom_psych_diag, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~  biomom_psych_diag + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_biomom_psych_t0_covar <- summary(lm(gc_exposure_score ~  biomom_psych_diag + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_biomom_psych_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Maternal psychiatric diagnosis", 
                                        "T0", 
                                        model_gc_biomom_psych_t0_covar$coefficients["biomom_psych_diag", "Estimate"],
                                        model_gc_biomom_psych_t0_covar$coefficients["biomom_psych_diag", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$biomom_psych_diag, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ biomom_psych_diag + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_biomom_psych_t2_covar <- summary(lm(gc_exposure_score ~ biomom_psych_diag + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_biomom_psych_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Maternal psychiatric diagnosis", 
                                        "T2", 
                                        model_gc_biomom_psych_t2_covar$coefficients["biomom_psych_diag", "Estimate"],
                                        model_gc_biomom_psych_t2_covar$coefficients["biomom_psych_diag", "Pr(>|t|)"],
                                        rsqpart)
```

## GC exposure and pregnancy complications

### T0

```{r}
# get distributions
table(all_t0$preg_complications, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ preg_complications + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_pregcomp_t0_covar <- summary(lm(gc_exposure_score ~ preg_complications + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_pregcomp_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Pregnancy complications", 
                                        "T0", 
                                        model_gc_pregcomp_t0_covar$coefficients["preg_complications", "Estimate"],
                                        model_gc_pregcomp_t0_covar$coefficients["preg_complications", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$preg_complications, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ preg_complications + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_pregcomp_t2_covar <- summary(lm(gc_exposure_score ~ preg_complications + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_pregcomp_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Pregnancy complications", 
                                        "T2", 
                                        model_gc_pregcomp_t2_covar$coefficients["preg_complications", "Estimate"],
                                        model_gc_pregcomp_t2_covar$coefficients["preg_complications", "Pr(>|t|)"],
                                        rsqpart)

```

## GC exposure and pre-term birth

### T0

```{r}
# get distributions
table(all_t0$preg_preterm, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ preg_preterm + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_preterm_t0_covar <- summary(lm(gc_exposure_score ~ preg_preterm + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_preterm_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Preterm birth", 
                                        "T0", 
                                        model_gc_preterm_t0_covar$coefficients["preg_preterm", "Estimate"],
                                        model_gc_preterm_t0_covar$coefficients["preg_preterm", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$preg_preterm, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ preg_preterm + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_preterm_t2_covar <- summary(lm(gc_exposure_score ~ preg_preterm + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_preterm_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Preterm birth", 
                                        "T2", 
                                        model_gc_preterm_t2_covar$coefficients["preg_preterm", "Estimate"],
                                        model_gc_preterm_t2_covar$coefficients["preg_preterm", "Pr(>|t|)"],
                                        rsqpart)

```


## GC exposure and maltreatment (yes/no)

### T0

```{r}
# get distributions
table(all_t0$Case.Control, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ Case.Control + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_malt_t0_covar <- summary(lm(gc_exposure_score ~ Case.Control + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_malt_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Maltreatment", 
                                        "T0", 
                                        model_gc_malt_t0_covar$coefficients["Case.Control", "Estimate"],
                                        model_gc_malt_t0_covar$coefficients["Case.Control", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$Case.Control, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ Case.Control + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_malt_t2_covar <- summary(lm(gc_exposure_score ~ Case.Control + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_malt_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Maltreatment", 
                                        "T2", 
                                        model_gc_malt_t2_covar$coefficients["Case.Control", "Estimate"],
                                        model_gc_malt_t2_covar$coefficients["Case.Control", "Pr(>|t|)"],
                                        rsqpart)

```

## GC exposure and psychopathology (yes/no)

### T0

```{r}
# get distributions
table(all_t0$PAPA_any, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ PAPA_any + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_psych_t0_covar <- summary(lm(gc_exposure_score ~ PAPA_any + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_psych_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Psychiatric diagnosis", 
                                        "T0", 
                                        model_gc_psych_t0_covar$coefficients["PAPA_any", "Estimate"],
                                        model_gc_psych_t0_covar$coefficients["PAPA_any", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$PAPA_any, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ PAPA_any + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_psych_t2_covar <- summary(lm(gc_exposure_score ~ PAPA_any + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_psych_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Psychiatric diagnosis", 
                                        "T2", 
                                        model_gc_psych_t2_covar$coefficients["PAPA_any", "Estimate"],
                                        model_gc_psych_t2_covar$coefficients["PAPA_any", "Pr(>|t|)"],
                                        rsqpart)

```

## GC exposure and stressful life events

### T0

```{r}
# get distributions
table(all_t0$life_events_total, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE,lm(gc_exposure_score ~ life_events_total + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))$partial.rsq[1]
model_gc_life_total_t0_covar <- summary(lm(gc_exposure_score ~ life_events_total + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t0, na.action = "na.omit"))
model_gc_life_total_t0_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Stressful life events", 
                                        "T0", 
                                        model_gc_life_total_t0_covar$coefficients["life_events_total", "Estimate"],
                                        model_gc_life_total_t0_covar$coefficients["life_events_total", "Pr(>|t|)"],
                                        rsqpart)

```

### T2

```{r}
# get distributions
table(all_t2$life_events_total, useNA = "ifany")

# fit model with covariates
rsqpart <- rsq.partial(adj = TRUE, lm(gc_exposure_score ~ life_events_total + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))$partial.rsq[1]
model_gc_life_total_t2_covar <- summary(lm(gc_exposure_score ~ life_events_total + age + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + smoke_exposure_score + alcohol_exposure_score, data = all_t2, na.action = "na.omit"))
model_gc_life_total_t2_covar

# append values to table
sumtable_gc[nrow(sumtable_gc) + 1,] <- c("Stressful life events", 
                                        "T2", 
                                        model_gc_life_total_t2_covar$coefficients["life_events_total", "Estimate"],
                                        model_gc_life_total_t2_covar$coefficients["life_events_total", "Pr(>|t|)"],
                                        rsqpart)

```


# Finalize summary table

### Correct p-values for multiple testing

```{r}
sumtable_gc$p_adjust <- p.adjust(p = sumtable_gc$p, method = "fdr")

```


### Export final table

```{r}
# make sure values are numeric and round
sumtable_gc <- sumtable_gc %>%
  mutate(across(c("coef", "p", "p_adjust", "adj_partial_Rsquared"), as.numeric),
         across(c("coef", "adj_partial_Rsquared"), round, 2),
         across(c("p", "p_adjust"), round, 3)) %>%
  relocate(adj_partial_Rsquared, .before = p)

# export
write.table(format(sumtable_gc, drop0trailing=FALSE), "../03_results/gc_exposure/gc_associations.txt", row.names = FALSE, sep = "\t")

```
