---
title: "Correlation Structure"
author: "Vera Natalia Karlbauer"
contact: vera_karlbauer@psych.mpg.de
date: "2024-08-02"
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 10 for manuscript on Drivers of FKBP5 DNA methylation.
- Prepares covariates for correlation analysis.
- Calculates calculation matrices between exposure scores and biodata that are either adjusted or unadjusted for cell types.
- Creates correlation plots (figure 5a and 5b).

# Setup

## Step 1: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(corrplot)

```

## Step 2: Load data
```{r}
# set temporary wd for chunk to folder with preprocessed data
setwd("../../01_data/02_proc")

# all_time = methylation & covariates for T0 & T2
load("covariates_both_timepoints.rda")

```


## Step 3: Prepare data for correlation plot of cell types & exposures

```{r}
# convert from long to wide format
biodata_correlations <- tidyr::pivot_wider(data = covariates_both_timepoints, id_cols = Subject, names_from = Time , values_from = c("CD34", "CD14", "Buccal", "smoke_exposure_score", "alcohol_exposure_score", "gc_exposure_score", "CRP_Ln", "Cort_AUCg"), names_sep = "_t")
# ungroup
biodata_correlations <- biodata_correlations %>%
  ungroup

# set to numeric
biodata_correlations <- mutate_all(biodata_correlations, as.numeric)

# create dataframe with renamed variables for easy plotting
biodata_corrplot <- select(biodata_correlations, -Subject)
biodata_corrplot <- biodata_corrplot %>%
  relocate(c(gc_exposure_score_t0, gc_exposure_score_t2, alcohol_exposure_score_t0,  alcohol_exposure_score_t2, smoke_exposure_score_t0, smoke_exposure_score_t2), .before = "CD34_t0") 
biodata_corrplot <- biodata_corrplot %>%
  dplyr::rename("CD34 cells T0" = CD34_t0,
                "CD34 cells T2" = CD34_t2,
                "CD14 cells T0" = CD14_t0, 
                "CD14 cells T2" = CD14_t2, 
                "Buccal cells T0" = Buccal_t0, 
                "Buccal cells T2" = Buccal_t2, 
                "Smoke exposure T0" = smoke_exposure_score_t0,
                "Smoke exposure T2" = smoke_exposure_score_t2,
                "Alcohol exposure T0" = alcohol_exposure_score_t0,
                "Alcohol exposure T2" = alcohol_exposure_score_t2,
                "GC exposure T0" = gc_exposure_score_t0,
                "GC exposure T2" = gc_exposure_score_t2,
                "CRP T0" = CRP_Ln_t0,
                "CRP T2" = CRP_Ln_t2,
                "Cortisol T0" = Cort_AUCg_t0,
                "Cortisol T2" = Cort_AUCg_t2)

```

## Step 4: Prepare data for partial correlations of exposures between timepoints adjusted for cell counts

```{r}
### residualize cell counts out of exposure variables and biodata

# create empty data frame for residuals
biodata_corrected_cellcounts <- data.frame(matrix(NA,
                          nrow = 115,
                          ncol = 0))

## glucocorticoid exposure
# t0
resid_gc_exposure_t0 <- lm(formula = gc_exposure_score_t0 ~ CD34_t0 + CD14_t0 + Buccal_t0, data = biodata_correlations)
# append
biodata_corrected_cellcounts$gc_exposure_score_t0 <- resid_gc_exposure_t0$residuals
# t2
resid_gc_exposure_t2 <- lm(formula = gc_exposure_score_t2 ~ CD34_t2 + CD14_t2 + Buccal_t2, data = biodata_correlations)
# append
biodata_corrected_cellcounts$gc_exposure_score_t2 <- resid_gc_exposure_t2$residuals

## alcohol exposure
# t0
resid_alcohol_exposure_t0 <- lm(formula = alcohol_exposure_score_t0 ~ CD34_t0 + CD14_t0 + Buccal_t0, data = biodata_correlations)
# append
biodata_corrected_cellcounts$alcohol_exposure_score_t0 <- resid_alcohol_exposure_t0$residuals
# t2
resid_alcohol_exposure_t2 <- lm(formula = alcohol_exposure_score_t2 ~ CD34_t2 + CD14_t2 + Buccal_t2, data = biodata_correlations)
# append
biodata_corrected_cellcounts$alcohol_exposure_score_t2 <- resid_alcohol_exposure_t2$residuals

## smoke exposure
# t0
resid_smoke_exposure_t0 <- lm(formula = smoke_exposure_score_t0 ~ CD34_t0 + CD14_t0 + Buccal_t0, data = biodata_correlations)
# append
biodata_corrected_cellcounts$smoke_exposure_score_t0 <- resid_smoke_exposure_t0$residuals
# t2
resid_smoke_exposure_t2 <- lm(formula = smoke_exposure_score_t2 ~ CD34_t2 + CD14_t2 + Buccal_t2, data = biodata_correlations)
# append
biodata_corrected_cellcounts$smoke_exposure_score_t2 <- resid_smoke_exposure_t2$residuals

# rename variables for easy plotting
biodata_corrected_cellcounts <- biodata_corrected_cellcounts %>%
  dplyr::rename("Smoke exposure T0" = smoke_exposure_score_t0,
                "Smoke exposure T2" = smoke_exposure_score_t2,
                "Alcohol exposure T0" = alcohol_exposure_score_t0,
                "Alcohol exposure T2" = alcohol_exposure_score_t2,
                "GC exposure T0" = gc_exposure_score_t0,
                "GC exposure T2" = gc_exposure_score_t2
                )
# make sure that everything is numeric
biodata_corrected_cellcounts <- mutate_all(biodata_corrected_cellcounts, as.numeric)

```

# Analysis 

## Step 1: Correlation plot of cell types, exposures, and biodata

```{r}
# create correlation matrix
corrmat_biodata = cor(biodata_corrplot, use = "na.or.complete", method = "spearman")
# test for significance
pmat_biodata = cor.mtest(as.matrix(biodata_corrplot), conf.level = 0.95, method = "spearman", alternative = "two.sided", exact = FALSE)

# initialize graphics device
tiff(file = "../04_figures/figure_5a_corrplot_celltypes_exposures_biodata.png", height = 7, width = 10, units = 'in', res = 700)

# plot
corrplot(corr = corrmat_biodata,
         method = "shade",
         type = "lower",
         addCoef.col = "black",
         tl.col = "black",
         tl.cex = 0.9,
         number.cex = 0.7,
         cl.cex = 0.9,
         p.mat = pmat_biodata$p,
         sig.level = 0.05,
         insig = "label_sig",
         pch = "    *",
         pch.col = "orange",
         pch.cex = 1.2)

# shut off graphics device
dev.off()
```

### Step 2: Partial correlations of exposures between timepoints adjusted for cell counts

```{r}
# create correlation matrix
corrmat_biodata = cor(biodata_corrected_cellcounts, use = "na.or.complete", method = "spearman")
# calculate p-values
pmat_biodata = cor.mtest(as.matrix(biodata_corrected_cellcounts), conf.level = 0.95, method = "spearman", alternative = "two.sided")

# initialize graphics device
tiff(file = "../04_figures/figure_5b_corrplot_adjusted_exposures.png", height = 7, width = 10, units = 'in', res = 700)

# plot
corrplot(corr = corrmat_biodata,
         method = "shade",
         type = "lower",
         addCoef.col = "black",
         tl.col = "black",
         tl.cex = 0.9,
         number.cex = 0.7,
         cl.cex = 0.9,
         p.mat = pmat_biodata$p,
         sig.level = 0.05,
         insig = "label_sig",
         pch = "     *",
         pch.col = "orange",
         pch.cex = 1.2)

# shut off graphics device
dev.off()

```

