---
title: 'Appendix A: Exploration of correlation between cell type estimations'
author: "Vera N. Karlbauer"
contact: vera_karlbauer@psych.mpg.de
date: "2024-09-30"
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Purpose:
- Script 11 for manuscript on Drivers of FKBP5 DNA methylation.
- Explores accuracy of Smith et al. (2014) method for saliva cell type estimation.  
- Runs cell type estimation based on Houseman algorithm and Zheng reference panel for saliva cell type estimation in BerlinLCS sample and compares resulting cell type proportions with original estimation.   
- Creates supplementary figure A1.
- Original cell type estimation based on Smith et al. (204), DOI: 10.1002/ajmg.b.32278 
- Zheng epithelial reference panel: Zheng et al. (2018), DOI: 10.2217/epi-2018-0037

# Setup

## Step 1: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(tibble)
library(tidyr)
library(BiocManager)
library(EpiDISH)
library(corrplot)

```


## Step 2: Load data
```{r}
# BerlinLCS pre-processed EPICs
load("../../01_data/01_raw/funnorm_npc4_2b_beta_values_nodupl.rda")

# original cell types
load("../../01_data/02_proc/all_time.rda")

```

## Step 3: Pre-process data

```{r}
# rename Smith-based cell types
all_time <- all_time %>%
  rename(CD14_Smith_Lowe = CD14,
         CD34_Smith_Lowe = CD34,
         Buccal_Smith_Lowe = Buccal)
```

# Cell type estimation with epithelial tissue reference panel (epiDISH)

## Estimate cell types

```{r}
# load reference
data(centEpiFibIC.m)

# estimate cell types
celltypes_saliva_zheng <- epidish(beta.m = beta_value, ref.m = centEpiFibIC.m, method = "RPC")$estF

# rename columns
celltypes_saliva_zheng <- as.data.frame(celltypes_saliva_zheng) %>%
  rownames_to_column(var = "MethID") %>%
  dplyr::rename(Epithelial_Zheng = Epi,
         Fibroblasts_Zheng = Fib,
         Immune_Zheng = IC)
# append to BerlinLCS dataframe
all_time <- left_join(all_time, celltypes_saliva_zheng, by = "MethID")

```

## Examine proportions

```{r}
all_time %>%
  group_by(Time) %>%
  summarise(mean_epithelial = mean(Epithelial_Zheng), mean_immune = mean(Immune_Zheng), mean_fibroblasts = mean(Fibroblasts_Zheng)) %>%
  round(digits = 4)

```

## Correlate proportions with Smith reference panel

```{r}
## Buccal (Smith) ~ Epithelial (Zheng)
# T0
cor.test(filter(all_time, Time == 0)$Buccal_Smith_Lowe, filter(all_time, Time == 0)$Epithelial_Zheng, method = "spearman")
# T2
cor.test(filter(all_time, Time == 2)$Buccal_Smith_Lowe, filter(all_time, Time == 2)$Epithelial_Zheng, method = "spearman")

## CD34 (Smith) ~ Immune (Zheng)
# T0
cor.test(filter(all_time, Time == 0)$CD34_Smith_Lowe, filter(all_time, Time == 0)$Immune_Zheng, method = "spearman")
# T2
cor.test(filter(all_time, Time == 2)$CD34_Smith_Lowe, filter(all_time, Time == 2)$Immune_Zheng, method = "spearman")

```

## Visualize correlations

```{r}
# filter data
celltypes_all <- pivot_wider(all_time,
                               id_cols = Subject,
                               names_from = Time,
                               values_from = c(Buccal_Smith_Lowe, CD34_Smith_Lowe,
                                               Epithelial_Zheng, Immune_Zheng),
                               names_sep = "_T") %>%
  select(-Subject)

# create correlation 
corrmat_zheng = cor(celltypes_all, use = "na.or.complete", method = "spearman")
corrmat_zheng <- corrmat_zheng[c("Buccal_Smith_Lowe_T0", "CD34_Smith_Lowe_T0", 
                                 "Buccal_Smith_Lowe_T2", "CD34_Smith_Lowe_T2"), 
                               c("Epithelial_Zheng_T0", "Immune_Zheng_T0", "Epithelial_Zheng_T2", "Immune_Zheng_T2")]
# calculate p-values
pmat_zheng = cor.mtest(as.matrix(celltypes_all), conf.level = 0.95, method = "spearman", alternative = "two.sided")
pmat_zheng <- pmat_zheng$p[c("Buccal_Smith_Lowe_T0", "CD34_Smith_Lowe_T0", 
                                 "Buccal_Smith_Lowe_T2", "CD34_Smith_Lowe_T2"), 
                               c("Epithelial_Zheng_T0", "Immune_Zheng_T0", "Epithelial_Zheng_T2", "Immune_Zheng_T2")]

# initialize graphics device
tiff(file = "../04_figures/figure_a1_saliva_celltypes_corrplot.png", height = 7, width = 7, units = 'in', res = 700)

# plot
corrplot(corr = corrmat_zheng,
         method = "shade",
         type = "full",
         addCoef.col = "black",
         tl.col = "black",
         tl.cex = 0.7,
         number.cex = 0.7,
         cl.cex = 0.7,
         cl.pos = "b",
         p.mat = pmat_zheng,
         sig.level = 0.05,
         insig = "label_sig",
         pch = "     *",
         pch.col = "orange",
         pch.cex = 2,
         mar = c(0, 0, 1, 0))

# shut off graphics device
dev.off()

```

