---
title: 'Appendix C: Maltreatment effects controlled for SES'
author: "Vera N. Karlbauer"
date: "2024-09-24"
contact: vera_karlbauer@psych.mpg.de
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 12 for manuscript on Drivers of FKBP5 DNA methylation. 
- Repeat all analyses from script 05 (effects of childhood maltreatment on FKBP5 DNA methylation) but correct for SES as additional covariate.
- Analyzes effects of maltreatment on FKBP5 methylation via linear regression. 
- Runs linear models for T0, T2, and over time including time and subject effects.  
- Exports results and publication tables (supplementary table C1).  
- The functions looping the linear model over 49 CpGs, specifying all relevant covariates, and providing a dataframe with adjusted R squared, regression coefficients, p-values, and fdr-corrected p-values are specified in the script 01_functions/02_appendix_c_functions_fkbp5.R .  
- All analyses are corrected for age, sex, genetic components (C1, C2, C3), cell type counts (CD34, CD14, buccal), prenatal smoke exposure, prenatal alcohol exposure, and SES. 

# Setup

## Step 0: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(lme4)
library(stringr)
library(rsq)
library(finalfit)
library(writexl)
library(haven)
library(labelled)

```


## Step 1: Source auxiliary functions

```{r}
source("../01_functions/02_appendix_c_functions_fkbp5.R")

```


## Step 2: Load data

```{r}
# set temporary wd for chunk to folder with preprocessed data
setwd("../../01_data/02_proc")

# all_t0 = methylation & covariates for T0
load("all_t0.rda")

# all_t2 = methylation & covariates for T2
load("all_t2.rda")

# all_time = methylation & covariates for T0 & T2
load("all_time.rda")

# additional phenotypes (includes SES)
psych_pheno <- read_sav("../01_raw/LCS_T0T4_TP2_Feb2019_fullversion_long-2.sav")

```
## Step 3: Process data

```{r}
## phenotypes
# select relevant variables
psych_pheno <- select(psych_pheno,
                      SUBJECT,
                      VISIT,
                      SES
                      )
# delete labels
psych_pheno <- labelled::remove_labels(psych_pheno)
psych_pheno <- labelled::remove_val_labels(psych_pheno)

# subject as character
psych_pheno$SUBJECT <- as.character(psych_pheno$SUBJECT)

```


### Step 4: Create joint dataframe

```{r}
# join dataframes
all_t0 <- left_join(all_t0, filter(psych_pheno, VISIT == 0), by = c("Subject" = "SUBJECT"))
all_t2 <- left_join(all_t2, filter(psych_pheno, VISIT == 2), by = c("Subject" = "SUBJECT"))
all_time <- left_join(all_time, psych_pheno, by = c("Subject" = "SUBJECT", "Time" = "VISIT"))

```


# Analysis

## SES-corrected maltreatment effects at T0

### Run statistical analysis

```{r}
# loop linear model with pre-specified covariates and maltreatment as main predictor over 49 CpGs
diffmeth_ses_malt_t0 <- ses_loop_lm_simple(data = all_t0, predictor_name = "Case.Control")

# count number of uncorrected significant hits
count(diffmeth_ses_malt_t0, sig_uncorrected)

# count number of fdr-corrected significant hits
count(diffmeth_ses_malt_t0, sig_corrected)

## export results
# round values
diffmeth_ses_malt_t0 <- diffmeth_ses_malt_t0 %>%
  mutate(across(c("adj.R", "partial_adj.R", "coef"), round, 2),
         # round p-values
  p = case_when(p >= 0.001 ~ as.character(round_tidy(p, 3)),
                p < 0.001 & p >= 0.0001 ~ as.character(format(round_tidy(p, 4), scientific = FALSE)),
                TRUE ~ "<0.001"),
  p_adjust = case_when(p_adjust >= 0.001 ~ as.character(round_tidy(p_adjust, 3)),
                p_adjust < 0.001 & p_adjust >= 0.0001 ~ as.character(format(round(p_adjust, 4), scientific = FALSE)),
                TRUE ~ "<0.001"))
# export
write.table(diffmeth_ses_malt_t0, "../03_results/maltreatment_ses/results_diffmeth_ses_malt_t0.txt", row.names = FALSE, sep = "\t")

```

### Prepare results table for publication

```{r}
# create new table with significant hits
pubtable_diffmeth_ses_malt_t0 <- diffmeth_ses_malt_t0 %>%
  filter(sig_uncorrected == "yes")

# add PCR name, position, timepoint and functional region
pubtable_diffmeth_ses_malt_t0 <- mutate(pubtable_diffmeth_ses_malt_t0,
                                 PCR = str_extract(pubtable_diffmeth_ses_malt_t0$CpG, ".*(?=_3)"),
                                 position = as.numeric(str_extract(pubtable_diffmeth_ses_malt_t0$CpG , "([3]{1}[0-9]{7})")),
                                 functional_region = case_when(
                                   PCR == "CTCF_PCR_17" ~ "3'TAD",
                                   PCR == "CTCF_PCR_5.1" ~ "5'TAD",
                                   PCR == "FKBP5_PCR_12" | PCR == "FKBP5_PCR_16" | PCR == "FKBP5_PCR_17" | PCR == "FKBP5_PCR_18" ~ "Proximal Enhancer",
                                   PCR == "FKBP5_int5_PCR_2" | PCR == "FKBP5_int5_PCR_3" ~ "Intron 5 Enhancer",
                                   PCR == "FKBP5_int7_PCR_1.2" | PCR == "FKBP5_int7_PCR_1.4" ~ "Intron 7 Enhancer",
                                   TRUE ~ NA_character_),
                                 timepoint = rep("T0", nrow(pubtable_diffmeth_ses_malt_t0))
                                 )

# select relevant variables
pubtable_diffmeth_ses_malt_t0 <- select(pubtable_diffmeth_ses_malt_t0,
                                    functional_region, PCR, position, timepoint, adj.R, coef, partial_adj.R, p, p_adjust)

# export
write_xlsx(pubtable_diffmeth_ses_malt_t0, "../05_tables/maltreatment_ses/pubtable_results_diffmeth_ses_malt_t0.xlsx", col_names = TRUE)

```

## Maltreatment effects at T2

### Run statistical analysis

```{r}
# loop linear model with pre-specified covariates and maltreatment as main predictor over 49 CpGs
diffmeth_ses_malt_t2 <- ses_loop_lm_simple(data = all_t2, predictor_name = "Case.Control")

# count number of uncorrected significant hits
count(diffmeth_ses_malt_t2, sig_uncorrected)

# count number of fdr-corrected significant hits
count(diffmeth_ses_malt_t2, sig_corrected)

## export results
# round values
diffmeth_ses_malt_t2 <- diffmeth_ses_malt_t2 %>%
  mutate(across(c("adj.R", "partial_adj.R", "coef"), round, 2),
         # round p-values
  p = case_when(p >= 0.001 ~ as.character(round_tidy(p, 3)),
                p < 0.001 & p >= 0.0001 ~ as.character(format(round_tidy(p, 4), scientific = FALSE)),
                TRUE ~ "<0.001"),
  p_adjust = case_when(p_adjust >= 0.001 ~ as.character(round_tidy(p_adjust, 3)),
                p_adjust < 0.001 & p_adjust >= 0.0001 ~ as.character(format(round(p_adjust, 4), scientific = FALSE)),
                TRUE ~ "<0.001"))

# export
write.table(diffmeth_ses_malt_t2, "../03_results/maltreatment_ses/results_diffmeth_ses_malt_t2.txt", row.names = FALSE, sep = "\t")

```

### Prepare results table for publication

```{r}
# create new table with significant hits
pubtable_diffmeth_ses_malt_t2 <- diffmeth_ses_malt_t2 %>%
  filter(sig_uncorrected == "yes")

# add PCR name, position, and functional region
pubtable_diffmeth_ses_malt_t2 <- mutate(pubtable_diffmeth_ses_malt_t2,
                                 PCR = str_extract(pubtable_diffmeth_ses_malt_t2$CpG, ".*(?=_3)"),
                                 position = as.numeric(str_extract(pubtable_diffmeth_ses_malt_t2$CpG , "([3]{1}[0-9]{7})")),
                                 functional_region = case_when(
                                   PCR == "CTCF_PCR_17" ~ "3'TAD",
                                   PCR == "CTCF_PCR_5.1" ~ "5'TAD",
                                   PCR == "FKBP5_PCR_12" | PCR == "FKBP5_PCR_16" | PCR == "FKBP5_PCR_17" | PCR == "FKBP5_PCR_18" ~ "Proximal Enhancer",
                                   PCR == "FKBP5_int5_PCR_2" | PCR == "FKBP5_int5_PCR_3" ~ "Intron 5 Enhancer",
                                   PCR == "FKBP5_int7_PCR_1.2" | PCR == "FKBP5_int7_PCR_1.4" ~ "Intron 7 Enhancer",
                                   TRUE ~ NA_character_),
                                 timepoint = rep("T2", nrow(pubtable_diffmeth_ses_malt_t2))
                                 )

# select relevant variables
pubtable_diffmeth_ses_malt_t2 <- select(pubtable_diffmeth_ses_malt_t2,
                                    functional_region, PCR, position, timepoint, adj.R, coef, partial_adj.R, p, p_adjust)

# export
write_xlsx(pubtable_diffmeth_ses_malt_t2, "../05_tables/maltreatment_ses/pubtable_results_diffmeth_ses_malt_t2.xlsx", col_names = TRUE)

```

## Maltreatment effects over time

### Run statistical analysis

```{r}
# loop linear model with pre-specified covariates and maltreatment as main predictor over 49 CpGs
diffmeth_ses_malt_time <- ses_loop_lm_time(data = all_time, predictor_name = "Case.Control")

# count number of uncorrected significant hits
count(diffmeth_ses_malt_time, sig_uncorrected)

# count number of fdr-corrected significant hits
count(diffmeth_ses_malt_time, sig_corrected)

## export results
# round values
diffmeth_ses_malt_time <- diffmeth_ses_malt_time %>%
  mutate(across(c("AIC", "p", "coef"), round, 2),
         # round p-values
  p = case_when(p >= 0.001 ~ as.character(round_tidy(p, 3)),
                p < 0.001 & p >= 0.0001 ~ as.character(format(round_tidy(p, 4), scientific = FALSE)),
                TRUE ~ "<0.001"),
  p_adjust = case_when(p_adjust >= 0.001 ~ as.character(round_tidy(p_adjust, 3)),
                p_adjust < 0.001 & p_adjust >= 0.0001 ~ as.character(format(round(p_adjust, 4), scientific = FALSE)),
                TRUE ~ "<0.001"))
# export
write.table(diffmeth_ses_malt_time, "../03_results/maltreatment_ses/results_diffmeth_ses_malt_time.txt", row.names = FALSE, sep = "\t")

```

### Prepare results table for publication

```{r}
# create new table with significant hits
pubtable_diffmeth_ses_malt_time <- diffmeth_ses_malt_time %>%
  filter(sig_uncorrected == "yes")

# add PCR name, position, and functional region
pubtable_diffmeth_ses_malt_time <- mutate(pubtable_diffmeth_ses_malt_time,
                                 PCR = str_extract(pubtable_diffmeth_ses_malt_time$CpG, ".*(?=_3)"),
                                 position = as.numeric(str_extract(pubtable_diffmeth_ses_malt_time$CpG , "([3]{1}[0-9]{7})")),
                                 functional_region = case_when(
                                   PCR == "CTCF_PCR_17" ~ "3'TAD",
                                   PCR == "CTCF_PCR_5.1" ~ "5'TAD",
                                   PCR == "FKBP5_PCR_12" | PCR == "FKBP5_PCR_16" | PCR == "FKBP5_PCR_17" | PCR == "FKBP5_PCR_18" ~ "Proximal Enhancer",
                                   PCR == "FKBP5_int5_PCR_2" | PCR == "FKBP5_int5_PCR_3" ~ "Intron 5 Enhancer",
                                   PCR == "FKBP5_int7_PCR_1.2" | PCR == "FKBP5_int7_PCR_1.4" ~ "Intron 7 Enhancer",
                                   TRUE ~ NA_character_),
                                 timepoint = rep("T0->T2", nrow(pubtable_diffmeth_ses_malt_time))
                                 )

# select relevant variables
pubtable_diffmeth_ses_malt_time <- select(pubtable_diffmeth_ses_malt_time,
                                    functional_region, PCR, position, timepoint, AIC, coef, p, p_adjust)

# export
write_xlsx(pubtable_diffmeth_ses_malt_time, "../05_tables/maltreatment_ses/pubtable_results_diffmeth_ses_malt_time.xlsx", col_names = TRUE)

```


