---
title: "Appendix D: Exploration of correlations between exposure scores and maltreatment"
author: "Vera N. Karlbauer"
contact: vera_karlbauer@psych.mpg.de
date: "2024-08-02"
output:
    html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose:
- Script 13 for manuscript on Drivers of FKBP5 DNA methylation. 
- Examine interactions between maltreatment and epigenetic prenatal alcohol exposure score on the epigenetic glucocorticoid exposure score at T0.  
- Statistically test for interaction effect in linear model, adjusted for the usual covariates age, sex, genetic components (C1, C2, C3), cell type counts (CD34, CD14, buccal) and prenatal smoke exposure.  
- Perform tercile split into high, medium, and low alcohol exposure and descriptively plot interaction between maltreatment and alcohol exposure on glucocorticoid exposure (supplementary figure D1).


# Setup

## Step 0: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(ggplot2)
library(fabricatr)

```


## Step 1: Import data

```{r}
# set temporary wd for chunk to folder with preprocessed data
setwd("../../01_data/02_proc")

# all_t0 = methylation & covariates for T0
load("all_t0.rda")

```


# Analysis

## Examine distribution of epigenetic prenatal alcohol exposure score

```{r}
hist(all_t0$alcohol_exposure_score)
```

## Test for interaction effect of maltreatment and alcohol exposure score on glucocorticoid exposure score

```{r}
# fit model
model_mt_alc_multiplicative <- lm(gc_exposure_score ~ Case.Control + Sex + C1 + C2 + C3 + CD34 + CD14 + Buccal + alcohol_exposure_score + smoke_exposure_score + Case.Control*alcohol_exposure_score, data = all_t0)
# inspect results
summary(model_mt_alc_multiplicative)
```


## Tercile split epigenetic prenatal alcohol exposure score into low, medium, and high exposure 

```{r}
all_t0 <- all_t0 %>%
  mutate(alcohol_cat = split_quantile(alcohol_exposure_score, 3))
all_t0$alcohol_cat <- factor(all_t0$alcohol_cat)
all_t0 <- all_t0 %>%
  mutate(alcohol_cat = case_match(alcohol_cat, "1" ~ "low", 
                                  "2" ~ "medium",
                                  "3" ~ "high",
                                  .default = NA_character_))

# crosstable for frequencies in each group
table(all_t0$Case.Control, all_t0$alcohol_cat)
```

## Plot and export epigenetic glucocorticoid exposure score at T0 stratified by maltreatment status and prenatal alcohol exposure

```{r}
ggplot(data = all_t0, 
       aes(x=factor(alcohol_cat, levels = c("low", "medium", "high"), label = c("low", "medium", "high")), y=gc_exposure_score, fill = factor(Case.Control))) +
  geom_boxplot() +
  scale_fill_manual("Maltreatment status",
                    values=c("#B2E5FF", "#51A3CC"), 
                    labels = c("non-maltreated", "maltreated")) +
  xlab("Epigenetic prenatal alcohol exposure score in terciles") +
  ylab("Epigenetic glucocorticoid exposure score") +
  # non-maltreated low alc = 31
  annotate(geom="text", x=0.8, y=2.3, label="N=31", color="black") +
  # maltreated low alc = 23
  annotate(geom="text", x=1.2, y=2.3, label="N=23", color="black") +
  # non-maltreated mid alc = 25
  annotate(geom="text", x=1.8, y=2.1, label="N=25", color="black") +
  # maltreated mid alc = 29
  annotate(geom="text", x=2.2, y=2.1, label="N=29", color="black") +
  # non-maltreated high alc = 23
  annotate(geom="text", x=2.8, y=1.8, label="N=23", color="black") +
  # maltreated high alc = 31
  annotate(geom="text", x=3.2, y=1.8, label="N=31", color="black") +
  theme_classic() +
  theme(plot.title = element_text(size=13))
ggsave(plot = last_plot(), filename = "supplementary_figure_a1_gc_maltreat_by_alc_T0.png", path = "../04_figures", device = 'png', height = 7, width = 10, dpi = 700)

```

